---
title: '6-3 Matching Methods - Solutions'
author: ""
date: " `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
header-includes:
   - \usepackage{amsmath}
---

```{r message=FALSE}

# libraries
xfun::pkg_attach2(c("tidyverse", # load all tidyverse packages
                    "here",      # set file path
                    "MatchIt",   # for matching
                    "optmatch",  # for matching
                    "cobalt"))   # for matching assessment


# chunk options ----------------------------------------------------------------
knitr::opts_chunk$set(
  warning = FALSE            # prevents warning from appearing after code chunk
)

# prevent scientific notation
# ----------
options(scipen = 999)

```

# Simulation 

Create simulation data. 


```{r}

# set seed
# ----------
set.seed(42) # set so that random process of generating data is reproducible 

# set the number of individuals for simulated dataset
# ----------
n = 1e4 # Number of individuals (smaller than last time)

# NOTE: Again, don't worry too much about how we're creating this dataset, 
# this is just an example.

# W3 scaled to have mu=24 and sigma=4 a la
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4789291/
# where k = mu^2/sigma^2 and theta = sigma^2/mu

# Also make treatment less likely so that there are more controls,
# and add ID column
df <- data.frame(ID = seq.int(n),
                 W1 = sample(0:2, size = n, replace = TRUE, 
                             prob = c(0.49,0.50,0.01)),
                 W2 = sample(0:5, size = n, replace = TRUE, 
                             prob = c(0.60,0.13,0.19,0.06, 0.015, 0.005)),
                 W3 = rgamma(n, 
                    shape = 36,
                    scale = (2/3)))
df <- df %>% 
  mutate(W3 = W3 + 8*(W1 == 1)+ 12*(W2==2) + 
           8*(W2==3) + 4*(W2==4) + (-4)*(W2 == 5),
         A = as.numeric(rbernoulli(n, 
                                   p = (0.16 + 0.07*(W1 > 0) + 0.21*(W2 == 0) - 
                                          0.1*(W3 > 25) ))),
         Y_0 = as.numeric(rbernoulli(n, 
                                     p = (0.87 + 0.035*(W1 > 0) + 0.05*(W2 > 0)) + 
                                       abs((W3 - 22)/100))),
         Y_1 = as.numeric(rbernoulli(n, 
                                     p = (0.34 + 0.035*(W1 > 0) + 0.3*(W2 > 0)) + 
                                       abs((W3 - 22)/100) + 0.2*(W3 > 30))),
         ITE = Y_1 - Y_0,
         Y_obs = as.numeric((A & Y_1) | (!A & Y_0))) 

ATE_true <- mean(df$ITE)
df_a1 <- df %>% filter(A == 1)
ATT_true <- mean(df_a1$ITE)

df <- df %>% select(-Y_0, -Y_1, -ITE)
df_a1 <- df_a1 %>% select(-Y_0, -Y_1, -ITE)
df_a0 <- df %>% filter(A == 0)
                 
head(df)
summary(df)
```


# Run Matching analysis for Exact matching

Run the match analysis to and store the details in an object called `match_exact_ate`.

```{r}

#
# ATE using matchit for exact
# --------------------------------------------------
match_exact_ate <- matchit(formula = A ~ W1 + W2, # formula (leaving out W3 bc it is continuous)
                           data = df,             # data 
                           method = "exact",      # specify method to use
                           estimand = "ATE")      # specify estimand you want

# view
summary(match_exact_ate)


```

There are a number of ways to view balance statistics. Check out the online documentation [here](https://kosukeimai.github.io/MatchIt/articles/assessing-balance.html#assessing-balance-with-matchit). However, if you want to pull out summary statistics so you can manually manipulate them to create you own table or ggplot, for example, you need the `pluck()` function to pull specific information for the lists (which is how they are stored). 

You can use the code `summary(match_exact_ate) %>% View()` to view the stored information which is saved as lists. 

```{r}
summary(match_exact_ate) %>% glimpse() #use View here instead
```

You can then `pluck()` elements of the lists as you see fit. Here's an example of how to "pluck"  the `sum.all` column, which you can then manipulate into a datframe and use in a ggplot or a table.  

```{r}



# pluck the "Summary of Balance for Matched Data" column
# note you will need to manually create a column to identify W1 and W2 rows 
summary_df <- 
  summary(match_exact_ate) %>% 
  pluck("sum.all") %>% 
  as_tibble() %>% 
  glimpse() 

```

Similarly, you can view summary statistics with another package called `cobalt`. 


```{r}

# summarize statistics 
bal.tab(match_exact_ate, un = TRUE, stats = c("m", "v", "ks"))


```

View the list of the information stored in this object. 

```{r}

# View list of the stored information 
bal.tab(match_exact_ate, un = TRUE, stats = c("m", "v", "ks")) %>% 
  glimpse() # use "View" here instead

```


Pluck the "balance" and create a data frame. 

```{r}

# pluck the "Balance" column
balance_statistics <- 
  bal.tab(match_exact_ate, un = FALSE, stats = c("m", "v", "ks")) %>% 
  pluck("Balance") %>% 
  glimpse()

```

